The ChEMBL database as Linked Open Data

### 2013-03-06 article.tex
Explained the design goals in more detail at the start of the Results section, and given more detail on the choice of some ontologies. But the fact is, there have not been that many design decisions otherwise.

We present here the updated ChEMBL-RDF, including a description of the mapping process 
and the identification of links to other datasets.

@@

For each of the common resource classes, a triple pattern was defined, following the
data available in the relational database. Figure~\ref{f1} shows how the various resource
classes are linked together, detailed in this section at a triple level.

The core concept in the ChEMBL database is that of the biological activity. 
While ChEMBL provides both the original activities as found in literature
and standardized values allowing comparison between studies, the triples only
make the latter available. The links to the literature are encoded using the CiTO ontology.
This results in a set of triples that look like:

@@

More than five thousand different activity types, represented by the chembl:type predicate,
are captured by the ChEMBL database.
The top five types are ``Potency'' (43\%), ``IC50'' (13\%), ``MIC'' (4.6\%), ``Inhibition'' (3.7\%),
and ``Ki'' (3.6\%). The activity types in ChEMBL-RDF are currently not available as, or
linked to, an ontology.

@@

The name and synonyms for each drug are
provided as labels. When InChI and InChIKeys are available for a drug, then these are
provided via the CHEMINF formalism (abbreviated here):

@@

The literature from which all this data was extracted, is replicated from the database,
but only a subset of the possible information about that literature is provided. In all cases 
the PubMed Identifiers (PMIDs) and Digital Object Identifiers (DOIs) are used to link
to authoritative database, as outlined later in this paper.
The properties are provided using the BIBO ontology:

@@

Two ontologies were recently developed that may address these issues for ChEMBL-RDF in future.
Firstly, the BioAssay Ontology, which addresses many of our
remaining concerns around making the experimental data machine readable~\cite{Visser2011}. For example,
it provides an ontology of activity types. Secondly,
the Quantity, Unit, Dimension and Type (QUDT) ontology which makes it possible
to express measurements in a standardised machine readable way~\cite{QUDT}. Both approaches are 
being explored in the Open PHACTS project and will be applied to ChEMBL-RDF as the Open PHACTS project progresses~\cite{Williams2012}.

>>>>

We present here the updated ChEMBL-RDF, including a description of the mapping process 
and the identification of links to other datasets. Because the primary purpose of of this
paper is to expose the ChEMBL data as linked data, less focus has been placed on ontologically
capturing the fine details of the pharmacological literature. This is the reason why
many concepts are not mapped to existing ontologies, though also the lack of ontologies
with a clear open license restrict the options.

@@

For each of the common resource classes, a triple pattern was defined, following the
data available in the relational database. Unless the ontological meaning of the data
in the database was very clear, the classes are annotated with concepts as implicitly
defined by the relational database. For example, the activity concept in the triples
merely follows the activity table in the database. As such, the concept of chembl:Activity
is implied by the content of that table in the relational database.

However, where the database content is well-defined, specific ontologies have been used.
For example, where the database specifies literature from which data was extraced,
the CiTO ontology is used. The choice of the ontologies is driven by the use cases and
compatibility with other Linked Data projects, rather than formal ontological arguments.
We have not observed ontological contradictions based on these choices.

Figure~\ref{f1} shows how the various resource classes are linked together, detailed in
this section at a triple level.
The core concept in the ChEMBL database is that of the biological activity.
While ChEMBL provides both the original activities as found in literature
and standardized values allowing comparison between studies, the triples only
make the latter available. The links to the literature are encoded using the CiTO ontology.
This results in a set of triples that look like:

@@

More than five thousand different activity types, represented by the chembl:type predicate,
are captured by the ChEMBL database.
The top five types are ``Potency'' (43\%), ``IC50'' (13\%), ``MIC'' (4.6\%), ``Inhibition'' (3.7\%),
and ``Ki'' (3.6\%). The activity types in ChEMBL-RDF are currently not available as, or
linked to, an ontology. The BioAssay Ontology could be a future option~\cite{Visser2011}.

@@

The name and synonyms for each drug are
provided as labels. If SMILES, InChI, or InChIKeys are given for a drug, then these are
provided via the CHEMINF formalism.
The CHEMINF ontology has been used here because of it's flexibility in defining
descriptor types and their origin and meaning. By allowing an ontology of descriptors,
we enable comparing values between databases semantically. Differences in, for example,
the SMILES may then be explained by the software used for that descriptor.

Example output for ChEMBL406142 looks like:

@@

The literature from which all this data was extracted, is replicated from the database,
but only a subset of the possible information about that literature is provided. In all cases 
the PubMed Identifiers (PMIDs) and Digital Object Identifiers (DOIs) are used to link
to authoritative database, as outlined later in this paper.
The properties are provided using the BIBO ontology, which is an open sourced ontology
for bibliographic information, compatible with the CiTO ontology:

@@

Two ontologies were recently developed that may address these issues for ChEMBL-RDF in future.
Firstly, the BioAssay Ontology, which addresses many of our
remaining concerns around making the experimental data machine readable. For example,
it provides an ontology of activity types. Secondly,
the Quantity, Unit, Dimension and Type (QUDT) ontology which makes it possible
to express measurements in a standardised machine readable way~\cite{QUDT}. Both approaches are 
being explored in the Open PHACTS project and will be applied to ChEMBL-RDF as the Open PHACTS project progresses~\cite{Williams2012}.
